Developing a Bubble Chamber Particle Discriminator Using Semi-Supervised Learning

### 2019-02-24 documentation/report.tex
Ok draft

\usepackage[margin=1in]{geometry}
\usepackage{amsmath}
\usepackage{biblatex}
\usepackage{graphicx}
\usepackage{float}
\usepackage{setspace}
\usepackage{fancyhdr}
\usepackage{lastpage}
\usepackage{titlesec}
\usepackage{enumitem}
\usepackage{subfig}
\usepackage{hyperref}
\usepackage{soul}

@@

\setlist[enumerate]{noitemsep, nolistsep}
\setlist[itemize]{noitemsep, nolistsep}

@@

\setlength{\belowcaptionskip}{-12pt}

\titleformat*{\section}{\large\bfseries}
\titleformat*{\subsection}{\normalsize\bfseries}
\titleformat*{\subsubsection}{\normalsize\bfseries}

\doublespacing

@@

\ul{Please note that the entire PICO collaboration are listed on the research whitepaper because of their contributions to the original PICO-60 dark matter experiment. They did not contribute to this study; I completed and documented it entirely independently.}

@@

For PICO-60, I developed and compared two sets of classification algorithms:

\begin{enumerate}
    \item I experimented further with supervised learning, exploring the following data formats to learn which was the best-performing solution:
    \begin{itemize}
        \item An 8-band Fourier transform of audio recorded in the bubble chamber. I applied a dense neural network with dropout \cite{dropout} and L2 regularization (which allowed for less overfitting to impure data).
        \item A full-resolution Fourier transform of the audio (all 50,001 data points). I once again applied a dense neural network with the Adam \cite{adam} optimizer.
        \item A raw audio waveform with a very deep 1D convolutional neural network (CNN) inspired by Dai et al.\ \cite{verydeepconvnets}.
        \item Images captured by cameras in the detector. I trained a 2D CNN on these, to learn whether they contain any relevant information.
    \end{itemize}
    \begin{figure}[ht]
        \centering
        \subfloat{\fbox{\includegraphics[width=0.9\textwidth]{audio_cropped}}}
        \qquad
        \subfloat{\fbox{\includegraphics[width=0.45\textwidth]{image_grid}}}
        \caption{Examples of the audio waveform and Fourier transform (above) and the images captured by cameras (below).}
    \end{figure}
    \item To even better handle impurities, I developed two novel semi-supervised learning algorithms. The impure alpha/WIMP-like labels were removed from a large portion of the training data, and my algorithms learned more accurate labels for this data.
    \begin{itemize}
        \item Iterative Cluster Nucleation is inspired by unsupervised clustering algorithms. A neural network is trained on a labeled set, and runs inference on the unlabeled set. Unlabeled examples with predictions closest to 0 or 1 are added to the labeled set; the newly synthesized labels are more accurate than the original labels.
        \item Gravitational Differentiation uses a novel piecewise exponential function to calculate final-layer gradients. Unlabeled examples ``gravitate'' toward more accurate predictions based on the confidence of the neural network's predictions.
    \end{itemize}
\end{enumerate}

@@

All models were trained on a training set, and a separate randomly selected validation set was used to select the best-performing technique and corresponding hyperparameters (with a grid search). A test set was used to verify the performance of the model ultimately chosen.

@@

\begin{itemize}
    \item The dense neural network, with banded Fourier transform data, increased validation accuracy to 97.3\% on validation data, up from the previous ML accuracy of 80.2\%.
    \item The full-resolution Fourier transform produced validation accuracy of 92.2\%.
    \item The raw waveform with a 1D very deep CNN reached 94.5\% accuracy.
    \item There was no useful information in the image data; only 63.0\% accuracy was achieved.
    \item Iterative cluster nucleation, using banded FT input data, improved accuracy to 99.1\%.
    \item Gravitational differentiation was even more accurate, at 99.7\%.
    \item The best gravitational differentiation configuration was not as accurate on final test data as on validation data, but was much better than supervised learning, at 98.3\%.
\end{itemize}

\begin{figure}[ht]
    \centering
    \fbox{\includegraphics[width=0.95\textwidth]{pico_final_results}}
    \caption{Validation accuracy, precision, recall, and loss for each technique in PICO-60. Error bars represent the 8th and 92nd percentiles.}
\end{figure}

@@

\begin{itemize}
    \item The dense neural network always produced 100\% false positives (worse than 91.0\%).
    \item The topological CNN produced a false positive rate of 92.6\% (also worse than 91.0\%).
    \item The cylindrical projection method with a 2D CNN improved the false positive rate from 91.0\% to 75.7\% (a 63.0\% reduction)!
\end{itemize}

\begin{figure}[ht]
    \centering
    \fbox{\includegraphics[width=0.6\textwidth]{deap_final_results}}
    \caption{Alpha background and WIMP removal rates for the highest-accuracy hyperparameter configuration of each technique in DEAP-3600.}
\end{figure}

@@

>>>>

\usepackage[margin=1in]{geometry}
\usepackage{amsmath}
\usepackage{biblatex}
\usepackage{parskip}
\usepackage{graphicx}
\usepackage{float}
\usepackage{fancyhdr}
\usepackage{lastpage}
\usepackage{titlesec}
\usepackage{enumitem}
\usepackage{subfig}
\usepackage{hyperref}
\usepackage{soul}

@@

@@

@@

\ul{Please note that the entire PICO collaboration are listed on the research whitepaper because of their contributions to the original PICO-60 dark matter experiment. They did not contribute to this study; I completed and documented it independently.}

@@

For PICO-60, I developed and compared two sets of classification algorithms: supervised learning, and semi-supervised learning.

\subsubsection{Supervised Learning}

I experimented further with supervised learning, exploring the following data formats to learn which was the best-performing solution:

\begin{enumerate}
    \item An 8-band Fourier transform of audio recorded in the bubble chamber. I applied a dense neural network with dropout \cite{dropout} and L2 regularization (which allowed for less overfitting to impure data).
    \item A full-resolution Fourier transform of the audio (all 50,001 data points). I once again applied a dense neural network with the Adam \cite{adam} optimizer.
    \item A raw audio waveform with a very deep 1D convolutional neural network (CNN) inspired by Dai et al.\ \cite{verydeepconvnets}.
    \item Images captured by cameras in the detector. I trained a 2D CNN on these, to learn whether they contain any relevant information.
\end{enumerate}

\begin{figure}[ht]
    \centering
    \subfloat{\fbox{\includegraphics[width=0.95\textwidth]{audio_cropped}}}
    \qquad
    \subfloat{\fbox{\includegraphics[width=0.45\textwidth]{image_grid}}}
    \caption{Examples of the audio waveform and Fourier transform (above) and the images captured by cameras (below).}
\end{figure}

\subsubsection{Semi-Supervised Learning}

To even better handle impurities, I developed two novel semi-supervised learning algorithms. The impure alpha/WIMP-like labels were removed from a large portion of the training data, and my algorithms learned more accurate labels for this data.

\begin{enumerate}
    \item Iterative Cluster Nucleation is inspired by unsupervised "clustering" algorithms, and takes advantage of the ``confidence'' of the neural network's predictions (that is, how close the predictions are to 0 or 1).

    First, a neural network is trained on a labeled set. After 30 epochs, it runs predictions on the unlabeled set. As the network trains, the confidence of these predictions increases. When $p < t$ or $p > 1 - t$ for a given prediction $p$ and a (small) confidence threshold $t$, the example is labeled correspondingly and added to the labeled set.

    It is hypothesized that accuracy will gradually \textit{improve} as the neural network's most confident predictions are fed back into the network as newly labeled training data.

    \item Gravitational Differentiation is an analog alternative to Iterative Cluster Nucleation. Using a novel piecewise exponential function for calculating final-layer derivatives, called $\mathrm{GravDiff}(p, \psi, g)$, unlabeled examples are caused to "gravitate" toward more accurate predictions based on the confidence of the neural network's predictions.

    The function is a transformation of the hyperbolic tangent that flattens the central range and comparatively exaggerates the asymptote on either side, defined as follows:
    \begin{center}
        $\mathrm{GravDiff}(p, \psi, g) = g \cdot \mathrm{sgn}(p) \cdot \lvert \mathrm{tanh}(2p - 1) \rvert ^ \psi$
    \end{center}
    where $p$ is the network's prediction, $\psi$ determines the rate at which the output approaches 0 as the confidence of the prediction decreases, and $g$ is the learning rate multiplier. The output is the final-layer gradient to be used in backpropagation.
\end{enumerate}

@@

Models were optimized using grid searches, in which every configuration within a hyperparameter space is trained and evaluated. All models were trained on a training set, and a separate randomly selected validation set was used to select the best-performing technique and corresponding hyperparameters.

Finally, the best-performing model was retrained on a newly selected training set, and a test set was used to verify its performance. In DEAP-3600, two test sets were used: one consisting of simulated data, and another consisting of 30 examples collected from the real-world detector (neutron calibration data, and background radiation).

@@

As seen in Figure \ref{pico_final_results}, the banded FFT input format was the most effective of all of the supervised learning models (including the previous study), reaching 97.3\% accuracy on validation data. The image data was shown to contain no meaningful information; it permitted only 63.0\% accuracy.

Semi-supervised learning produced better accuracy than supervised learning. Gravitational differentation was the best overall model, reaching 99.7\% accuracy on validation data and 98.3\% on test data. This confirms my hypothesis; it is indeed beneficial to train on the neural network's most confident predictions.

\begin{figure}[ht]
    \centering
    \fbox{\includegraphics[width=0.9\textwidth]{pico_final_results}}
    \caption{\label{pico_final_results} Validation accuracy, precision, recall, and loss for each technique in PICO-60. Error bars represent the 8th and 92nd percentiles.}
\end{figure}

@@

Both the dense neural network and topological CNN produced a higher rate of false positives than the conventional classifier. However, as seen in Figure \ref{deap_final_results}, the cylindrical projection method improved the false positive rate from 91.0\% to 75.7\% (a 63.0\% reduction)!

When tested on real-world data, the cylindrical projection method removed 73.9\% of nuclear recoil (neutron calibration) events (in line with the simulation).

However, only 91.2\% of background radiation events were removed. This indicates either that the simulation is not an accurate representation of the real-world detector, or (less probably) that WIMPs are present in the background data.

\begin{figure}[ht]
    \centering
    \fbox{\includegraphics[width=0.55\textwidth]{deap_final_results}}
    \caption{\label{deap_final_results} Background and WIMP removal rates for the highest-accuracy hyperparameter configuration of each technique in DEAP-3600.}
\end{figure}

@@

Additionally, the unusual predictions made on real-world data provide important diagnostic information on the DEAP-3600 simulation (assuming that WIMPs are not present).
